Making Sequential Programs Deterministic

### 2019-04-05 experiments.tex
Continued work on case study and experiments

The state of the art in flaky test detection is NonDex~\cite{nondex}.
\Cref{sec:related} explains how NonDex works. This section compares the
errors reported by NonDex and \theDeterminismChecker. We ran all our
experiments on a 2.3 GHz Intel Core i5 processor with a RAM configuration of 8 GB 2133 MHz.
\TheDeterminismChecker doesn't show any noticeable performance change at compile time.
\TheDeterminismChecker runs on any project that builds with Maven or Gradle and requires
Java version 8 or higher (support for annotations was added in Java 8).

@@

We ran NonDex on the same commit of Randoop as we used for the case study
of \cref{sec:randoop-case-study}.  This version contains 3 nondeterminism bugs that
\theDeterminismChecker found.

We modified Randoop by deleting tests that were skipped by its buildfile,
because the NonDex Gradle plugin does not respect those settings.
After that, NonDex ran without problems on Randoop.
It completed successfully without reporting any flaky tests.
We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
Bug 1 and Bug 2 have corresponding test cases whereas Bug 3 does not.
As NonDex is a \textit{flaky test} detection tool, Bug 3 is outside of its scope.
The reason for nondeterminism in Bug 2 was a call to the \<System.getProperty()> method which
is not modeled by NonDex.
This indicates that the additional programmer effort required by \TheDeterminismChecker is justified
as our tool discovers non-trivial issues and doesn't depend on code coverage or other runtime heuristics.

@@

\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
This section determines whether NonDex finds errors that \theDeterminismChecker does not.
Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
21 of them. We defer the reader to~\cite{nondex} for a complete list of these projects
and the details of commits hashes in which NonDex detected flaky tests.
The authors also reported the sources of flakiness
after manually inspecting these tests.

We ran \TheDeterminismChecker on all these benchmarks with the objective
of measuring the recall of our tool. We cloned all 21 repos and switched to the commit that had a flakiness issue as reported 
in~\cite{nondex}. Of these projects, \<handlebars> and \<oryx> have compile time failures which prevent us from running \TheDeterminismChecker on them. For benchmarks \<visualee>, \<easy-batch>, and \<vraptor>, we couldn't find
the sources of flakiness as reported in~\cite{nondex} anywhere in the respective repositories.
\TheDeterminismChecker reported the nondeterministic source locations correctly in the rest of the projects.
All these benchmarks and instructions to reproduce our results can be found at \href{https://github.com/t-rasmud/determinismCaseStudies}{https://github.com/t-rasmud/determinismCaseStudies}.

All the flakiness that NonDex found was due to 7 methods in 3 classes. We enumerate all the methods corresponding to
each class in \cref{fig:flaky-sources}.
\begin{figure}
    \begin{tabular}{lll}
        \<Class> & : \<getDeclaredFields>, \<getDeclaredMethods>, \<getFields> \\
        \<DateFormatSymbols> & : \<getZoneStrings> \\
        \<HashMap> & : \<entrySet>, \<keySet>, \<values>
    \end{tabular}
\caption{Souces of flakiness in~\cite{nondex}.}
\label{fig:flaky-sources}
\end{figure}

We provide samples of the annotated source code from these benchmarks and the errors as reported by \TheDeterminismChecker 
in \cref{fig:nondex-source}.
\begin{figure}
    \centering
    \begin{subfigure}[b]{0.95\textwidth}
        \begin{verbatim}
    static public FieldAccess get (Class type) {
        @Det ArrayList<@Det Field> fields = new ArrayList<Field>();
        ...
        while (nextClass != Object.class) {
            @Det Field @Det[] declaredFields = nextClass.getDeclaredFields();
            ...
        }
    }
    [ERROR] FieldAccess.java:[102,87] 
    incompatible types in assignment.
    found   : @Det Field @OrderNonDet []
    required: @Det Field @Det []
        \end{verbatim}
        \caption{Annotated source code from reflectasm. \vspace{0.5cm}}
        \label{code-reflectasm}
    \end{subfigure}

    \begin{subfigure}[b]{0.95\textwidth}
        \begin{verbatim}
    protected SimpleDataEvent createNextEvent() {
        for (@Det Entry<@Det String, @Det FieldType> entry : fields.entrySet()) {
            ...
        }
        ...
    }
    [ERROR] DataDictionaryEventSource.java:[73,78] 
    incompatible types in enhanced for loop.
    found   : @NonDet Entry<@Det String, @Det FieldType>
    required: @Det Entry<@Det String, @Det FieldType>
        \end{verbatim}
        \caption{Annotated source code from ActionGenerator.}
        \label{code-actiongenerator}
    \end{subfigure}    
    \caption{Errors detected by \TheDeterminismChecker.}
    \label{fig:nondex-source}
\end{figure}
As can be seen from these figures, annotating local arrays and collections
has been sufficient to discover the sources of nondeterminism in most of the projects.
The rest of the projects required annotating the return types of methods.

For the benchmark \<commons-lang>, \cite{nondex} report multiple failing tests (14 flaky tests) due to calls
to \<Class.getDeclaredFields>. Upon inspecting the source code, we find that all of invocations to
\<Class.getDeclaredFields> are made from 5 locations. This means that annotating 5 lines of source code
would have been sufficient to identify the nondeterminism. Having said that, we admit that there could be
significant programmer effort involved in annotating the whole program. On the other hand,~\cite{nondex}
state "we found that manually inspecting these failures was
rather challenging, and we leave it as future work to automate
debugging test failures due to ADINS code". \TheDeterminismChecker reports source locations
which makes it easier for the programmer to fix issues and is orthogonal to the approach taken by NonDex which
reports failing tests.

\TheDeterminismChecker is also capable of detecting nondetermism in test cases. We provide a stub file
for \<junit> in which we annotated the parameters of all the \<assert> methods as \<Det>.
This will cause \TheDeterminismChecker to report any \<junit> test that passes nondeterministic arguments to 
\<assert>s. The test case doesn't even have to be executed for out our tool to detect errors.
It took one of our authors under an hour to create the stub file for junit. This gives us
empirical evidence of the ease of extensibility of our tool.

@@

\section{Comparison to Deflaker}
Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. 
While we would have like to run DeFlaker on Randoop for a fair comparison of the tools,
we were unable to do so because Deflaker works with projects built with Maven and Randoop
uses Gradle for its build system. We plan to convert Randoop's build system to Maven and
perform this case study.
%Ideally, we would have 
%liked to run all three tools (NonDex, DeFlaker, and \TheDeterminismChecker) on an open source project
%that wasn't explored by any of the tool for a fair comparison. Unfortunately, there are issues with the
%DeFlaker plugin and we weren't able to run it on any project. We have created a Github issue for the same
%in its repository.
%\todo{Deflaker works now. Tried on retrofit: only works where there is sufficient change in code coverage between head and its parent.
%Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
%\todo{Deflaker requires Maven build. Randoop is on Gradle.}

>>>>

The state of the art in flaky test detection is NonDex~\cite{nondex}.
\Cref{sec:related} explains how NonDex works. This section compares the
errors reported by NonDex and \theDeterminismChecker.

@@

We ran NonDex on the same commit of Randoop as we used for the case study
of \cref{sec:randoop-case-study}.  This version contains all the nondeterminism bugs that
\theDeterminismChecker found.

We modified Randoop by deleting tests that were skipped by its buildfile,
because the NonDex Gradle plugin does not respect those settings.
After that, NonDex ran without problems on Randoop.
NonDex reported no flaky tests.
\todo{Revise the following text.
  I renumbered the bugs without realizing that this section refers to them
  by number, and I don't know which one these refer to any
  more.
  Please use cross-references or macros.  Also, this needs to be expanded
  to account for the additional bugs.}
We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
Bug 1 and Bug 2 have corresponding test cases whereas Bug 3 does not.
As NonDex is a \textit{flaky test} detection tool, Bug 3 is outside of its scope.
The reason for nondeterminism in Bug 2 was a call to the \<System.getProperty()> method which
is not modeled by NonDex.

@@

\begin{figure}
    \begin{tabular}{lll}
        \<Class> & : \<getDeclaredFields>, \<getDeclaredMethods>, \<getFields> \\
        \<DateFormatSymbols> & : \<getZoneStrings> \\
        \<HashMap> & : \<entrySet>, \<keySet>, \<values>
    \end{tabular}
\caption{Souces of flakiness known to NonDex~\cite{nondex}.}
\label{fig:flaky-sources}
\end{figure}

\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
This section determines whether NonDex finds errors that \theDeterminismChecker does not.
Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
21 of them~\cite{nondex}.
The authors also reported the sources of flakiness
after manually inspecting these tests. 
The flakiness that NonDex found was due to 7 methods in 3 classes (\cref{fig:flaky-sources}).

We tried to run \theDeterminismChecker on all these benchmarks, at the
commit given in the NonDex paper.
Some of the projects had moved, or did not build because their dependencies
had moved or were no longer available.  We repaired all these issues and
were able to compile all but two
projects, handlebars and oryx.  Some of the remaining projects did not pass
their tests, but that did not hinder us since \theDeterminismChecker works
at compile time.

For three of the projects, we could not find the flakiness reported in the
NonDex paper.
\todo{Important: give a sentence description of each one.}
\todo{visualee}
\todo{easy-batch}
\todo{vraptor}.

This left 16 projects.  We ran \theDeterminismChecker on the part of the
project that the NonDex authors determined as flaky.  In every case,
\theDeterminismChecker issued a warning on the nondeterministic code.  In
other words, \theDeterminismChecker's recall was 100\%.

All these benchmarks and instructions to reproduce our results can be found at \myurl{https://github.com/t-rasmud/determinismCaseStudies}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.95\textwidth}
        \begin{verbatim}
    static public FieldAccess get(Class type) {
        ...
        while (nextClass != Object.class) {
            Field[] declaredFields = nextClass.getDeclaredFields();
            ...
        }
    }
        \end{verbatim}
        \caption{Nondeterministic source code from reflectasm.}
        \label{code-reflectasm}
    \end{subfigure}

    \begin{subfigure}[b]{0.95\textwidth}
        \begin{verbatim}
    protected SimpleDataEvent createNextEvent() {
        for (Entry<String, FieldType> entry : fields.entrySet()) {
            ...
        }
        ...
    }
        \end{verbatim}
        \caption{Annotated source code from ActionGenerator.}
        \label{code-actiongenerator}
    \end{subfigure}    
    \caption{Errors detected by \TheDeterminismChecker.}
    \label{fig:nondex-source}
\end{figure}

\Cref{fig:nondex-source} show samples of nondeterministic code from these
benchmarks.
We annotated the benchmarks based on 
assumptions made downstream of the shown code.

In reflectasm (\cref{code-reflectasm}), we changed the type of field
\<declaredFields> from \<Field[]> to \<@Det Field @Det []>.  That type
means a deterministic array of deterministic \<Field>, analogously to
\codeid{@Det List<@Det Field>}.  Then, \theDeterminismChecker issued a
warning at the assignment, because \<getDeclaredMethods> returns \<@Det
Field @Det []>, which is an order-nondeterministic array of deterministic
\<Field>s.

The ActionGenerator code (\cref{code-actiongenerator}) is similar.  Other
code assumes that \<entry> is deterministic, but annotating it as \<@Det>
leads to a warning from \theDeterminismChecker that iterating over
\<fields.entrySet()> (which is itself \<@OrderNonDet>) yields \<@NonDet>
\<Entry> values..

Nondex found 14 flaky tests in Apache Commans Lang due to calls
to \<Class.getDeclaredFields>. 
There are only 5 invocations of \<Class.getDeclaredFields>, so annotating 5 lines of source code
would have been sufficient to identify all that nondeterminism.
Having said that, we admit that there could be
significant programmer effort involved in annotating the whole program. On
the other hand, the NonDex authors
state ``we found that manually inspecting these failures was
rather challenging, and we leave it as future work to automate
debugging test failures.''
\TheDeterminismChecker reports source locations
which makes it easier for the programmer to fix issues, and the annotation
effort serves as valuable documentation and prevents regressions.

\TheDeterminismChecker is capable of detecting nondetermism in test
cases, as some of these examples are.
We wrote specifications for the
JUnit library (in addition to the JDK and some other libraries).
The specifications are the \<@Det> type qualifier on the formal parameters
to the \<assert>* methods.
\TheDeterminismChecker reports any \<junit> test that passes nondeterministic arguments to
\<assert>*, even if the test case is not executed.

@@

\section{Comparison to Deflaker}
Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. 
We were unable to run DeFlaker on Randoop, because Deflaker works with projects built with Maven and Randoop
uses Gradle for its build system. We plan to convert Randoop's build system to Maven and
perform this case study.
%Ideally, we would have 
%liked to run all three tools (NonDex, DeFlaker, and \TheDeterminismChecker) on an open source project
%that wasn't explored by any of the tool for a fair comparison. Unfortunately, there are issues with the
%DeFlaker plugin and we weren't able to run it on any project. We have created a Github issue for the same
%in its repository.
%\todo{Deflaker works now. Tried on retrofit: only works where there is sufficient change in code coverage between head and its parent.
%Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
%\todo{Deflaker requires Maven build. Randoop is on Gradle.}

### 2019-04-05 randoop-case-study.tex
Continued work on case study and experiments

@@

@@

\caption{Fixes made by the Randoop developers in response to our bug report
  about use of environment variables.  Lines starting with ``\<->'' were
  removed and those starting with ``\<+>'' were added.}
\label{fig:randoop-bug-getenv}
\end{figure}

>>>>

%% This is not relevant in this section of the paper, where no times are given.
% We ran all our
% experiments on a 2.3 GHz Intel Core i5 processor with 8 GB of RAM.

@@

% run: ./gradlew clean compileJava;./gradlew clean compileJava;./gradlew clean compileJava;./gradlew clean compileJava;./gradlew clean compileJava
% then read the times reported by Gradle.
% javac 5 trials:   31, 19, 18, 18, 18.
% \theDeterminismChecker 5 trials: 45s 39s 32s 31s 31s
\<javac> takes 18 seconds to compile all files of Randoop.
While also running \theDeterminismChecker as a compiler plugin, \<javac> takes 32 seconds to verify determinism and compile all files of Randoop.
% godwit.  Info comes from `cat /proc/cpuinfo` and from `top`.
These numbers are the median of 5 trials on an 8-core Intel i7-3770 CPU running at 3.40GHz with 
32GB of memory.

@@

\caption{Fixes made by the Randoop developers in response to our bug report
  about use of environment variables.  Lines starting with ``\<->'' were
  removed and those starting with ``\<+>'' were added.
  \TheDeterminismChecker verified all other uses of system and Java properties.}
\label{fig:randoop-bug-getenv}
\end{figure}
